### 2024-04-05 docs/SOSP 2024/caching.tex
Add HIR MIR optimization section

One of the major changes we make to the original design of \Treebeard{} \cite{Treebeard} is to
separate the implementation of representations from the rest of the compiler. This allows
us to implement representations as plugins to the compiler. We do this by defining an interface 
that each representation needs to implement. The code generator is implemented using methods 
on the representation interface thus hiding details of the actual representation from the core 
compiler. The interface abstracts the following details. 
\begin{itemize}
  \item \textbf{Buffer allocation:} The representation object exposes methods that generate 
  buffer allocations in the IR. 
  \item \textbf{Moving to child nodes:} Given the value of the predicate at a node (or tile),
  generate code to move to the appropriate child node of the current node.
  \item \textbf{Details of leaves:} The interface abstracts determining whether the current
  node is a leaf (which is needed to terminate walks) and how to get the value of the leaf
  (which is needed to get a tree's prediction).
  \item \textbf{Caching trees:} Since reading the trees into shared memory on GPU (or prefetching)
  them on CPU require knowledge of the buffer layout, the task of generating caching code for trees
  is delegated to the representation object. 
  \item \textbf{Loading thresholds and feature indices:} The representation object provides MLIR 
  rewrite patterns to lower operations that load thresholds and feature indices to LLVM IR. This
  is necessary because the details of how to load the thresholds and feature indices from the model 
  buffers is representation-specific.
\end{itemize}

>>>>

One of the major changes we make to the original design of \Treebeard{} \cite{Treebeard} is to
separate the implementation of representations from the rest of the compiler. This allows
us to implement representations as plugins to the compiler. We do this by defining an interface 
that each representation needs to implement. The code generator is implemented using methods 
on the representation interface thus hiding details of the actual representation from the core 
compiler. The interface abstracts the following details. 
\begin{itemize}
  \item \textbf{Buffer allocation:} The representation object exposes methods that generate 
  buffer allocations in the IR. 
  \item \textbf{Moving to child nodes:} Given the value of the predicate at a node (or tile),
  generate code to move to the appropriate child node of the current node.
  \item \textbf{Leaf representation:} The interface abstracts determining whether the current
  node is a leaf (which is needed to terminate walks) and how to get the value of the leaf
  (which is needed to get a tree's prediction).
  \item \textbf{Caching trees:} Since reading the trees into shared memory on GPU (or prefetching)
  them on CPU require knowledge of the buffer layout, the task of generating caching code for trees
  is delegated to the representation object. 
  \item \textbf{Loading thresholds and feature indices:} The representation object provides MLIR 
  rewrite patterns to lower operations that load thresholds and feature indices to LLVM IR. This
  is necessary because the details of how to load the thresholds and feature indices from the model 
  buffers is representation-specific.
\end{itemize}

### 2024-04-05 docs/SOSP 2024/hir-mir-opts.tex
Add HIR MIR optimization section

\begin{itemize}
  \item The compiler performs a number of optimizations on the HIR and MIR to improve the performance of the generated code.
  \item Optimizations in HIR mainly transform the model itself. Tiling, tree reordering and padding are some of the 
  optimizations that are performed on the HIR. \TODO{Should we use the MICRO paper overview for this part?}
  \item Explain lowering to MIR using the schedule here.
  \item Optimizations on the MIR include tree walk unrolling, tree walk interleaving, loop tiling, parallelization etc.
\end{itemize}

>>>>

% \begin{itemize}
%   \item The compiler performs a number of optimizations on the HIR and MIR to improve the performance of the generated code.
%   \item Optimizations in HIR mainly transform the model itself. Tiling, tree reordering and padding are some of the 
%   optimizations that are performed on the HIR. \TODO{Should we use the MICRO paper overview for this part?}
%   \item Explain lowering to MIR using the schedule here.
%   \item Optimizations on the MIR include tree walk unrolling, tree walk interleaving, loop tiling, parallelization etc.
% \end{itemize}

The \Treebeard{} infrastructure was originally designed to target CPUs \cite{Treebeard}. 
However, it implements several optimizations on the HIR and MIR that 
can be leveraged across target processors and we find that these 
are beneficial for GPUs as well. This reuse of optimizations is possible 
because the intermediate representations on which these optimizations are performed 
are abstract and are designed to be target-independent. We briefly review theese 
optimizations below.

\subsection{Optimizations on High-Level IR}
We augment the existing \Treebeard{} infrastructure with loop rewrites on 
the HIR that are implemented through the scheduling language (Section \ref{sec:schedule}).
We use these to implement the automatic scheduling described in Section \ref{sec:exploring}. 
Additionally, the \Treebeard{} infrastructure implements HIR transformations to reorder and pad 
trees. It also implements tree tiling transformations on the HIR \cite{Treebeard}. We reuse the 
reordering and padding transformations on the HIR for GPUs. However, we found that 
tiling trees was not beneficial for GPUs. This is because the tiling transformations
introduces redundant computation inorder to vectorize computation on CPUs where 
all lanes need to follow the same control flow. However, on SIMT GPUs, we find that 
the benefits of tiling (coalescing memory accesses) do not outweigh the cost of
redundant computation. We leave an investigation of this for future work.

\subsection{Optimizations on Mid-Level IR}
The original \Treebeard{} infrastructure implements optimizations like 
tree-walk unrolling, tree-walk interleaving, and parallelization
on the MIR. These optimizations are beneficial for GPUs as well.
and the design of \Treebeard{} allows us to reuse the tree-walk 
unrolling and tree-walk interleaving optimizations on the MIR for GPUs.
\TODO{Should we talk about how interleaving is implemented as a 
statemachine and therefore it can be used across representations 
and tile traversal techniques?}

\subsection{A Note on Low-level IR}
Some changes to the original \Treebeard{} design were required to get
LIR to correctly lower to GPU code. The most important of these was 
the change to how the compiler implements support for in-memory 
representations of models (Section \ref{sec:representations}).
With these design changes, we were able to reuse much of the CPU 
implementation while customizing some parts for GPUs (for example,
buffers need to be allocated differently for CPU and GPU, caching 
is implemented differently etc.).
